---
title: "Introduction to PSPLINE"
author: "Yoon-Bae Jun"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to PSPLINE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

`PSPLINE` fits generalized partially linear single-index models of the form

\[
g\{E(Y_i\mid X_i,Z_i)\}
=
X_i^\top\beta+\phi(Z_i^\top\alpha),
\]

where \(X_i\) enters linearly, \(Z_i\) enters through a single index, and
\(\phi(\cdot)\) is an unknown smooth function estimated using penalized
splines. The package normalizes \(\alpha\) so that \(\|\alpha\|=1\) and its
first component is positive. The current implementation supports binomial and
Poisson outcomes.

# Installation

```{r installation, eval=FALSE}
remotes::install_github("junpeea/PSPLINE")
```

```{r load-package}
library(PSPLINE)
```

# A binomial example

## Simulate data

```{r simulate-binomial}
set.seed(2026)

n <- 250

Z <- cbind(
  z1 = runif(n, 0, 2),
  z2 = runif(n, 0, 2),
  z3 = runif(n, 0, 2)
)

X <- cbind(
  x1 = rnorm(n),
  x2 = rbinom(n, 1, 0.5)
)

alpha_true <- c(1, 1, 1) / sqrt(3)
beta_true <- c(0.8, -0.5)

u_true <- as.numeric(Z %*% alpha_true)
phi_true <- sin(2 * u_true)
phi_true <- phi_true - mean(phi_true)

eta_true <- as.numeric(X %*% beta_true + phi_true)
probability <- plogis(eta_true)
y <- rbinom(n, size = 1, prob = probability)

mean(y)
```

## Fit the model

```{r fit-binomial}
fit_binomial <- LuGPLSIM(
  y = y,
  X = X,
  Z = Z,
  family = binomial(),
  M = 4,
  maxit = 50,
  tol = 1e-6
)
```

```{r print-and-summary}
fit_binomial
summary(fit_binomial)
```

Check the identifiability constraint:

```{r alpha-constraint}
sum(fit_binomial$alpha_hat^2)
fit_binomial$alpha_hat[1]
```

# Prediction components

```{r prediction-components}
u_hat <- predict(
  fit_binomial,
  newX = X,
  newZ = Z,
  type = "index"
)

smooth_hat <- predict(
  fit_binomial,
  newX = X,
  newZ = Z,
  type = "smooth"
)

linear_hat <- predict(
  fit_binomial,
  newX = X,
  newZ = Z,
  type = "linear"
)

eta_hat <- predict(
  fit_binomial,
  newX = X,
  newZ = Z,
  type = "link"
)

probability_hat <- predict(
  fit_binomial,
  newX = X,
  newZ = Z,
  type = "response"
)

head(cbind(u_hat, smooth_hat, linear_hat, eta_hat, probability_hat))
range(probability_hat)
```

# Plot the nonlinear component

```{r plot-method}
plot(fit_binomial)
```

A direct plot can also be created from the prediction methods:

```{r custom-plot}
ordering <- order(u_hat)

plot(
  u_hat[ordering],
  smooth_hat[ordering],
  type = "l",
  xlab = expression(hat(u) == Z^T * hat(alpha)),
  ylab = expression(hat(phi)(hat(u))),
  main = "Estimated single-index smooth"
)
```

# Variance estimation

```{r vcov-model}
V_model <- vcov(
  fit_binomial,
  type = "model",
  component = "parameters"
)

dim(V_model)
round(V_model, 4)
```

Separate blocks are available:

```{r vcov-blocks}
V_alpha <- vcov(
  fit_binomial,
  type = "model",
  component = "alpha"
)

V_beta <- vcov(
  fit_binomial,
  type = "model",
  component = "beta"
)

dim(V_alpha)
dim(V_beta)
```

The sandwich covariance is obtained with:

```{r vcov-sandwich}
V_sandwich <- vcov(
  fit_binomial,
  type = "sandwich",
  component = "parameters"
)

dim(V_sandwich)
```

Leverage-adjusted and finite-sample-corrected covariance estimators are also
available:

```{r vcov-adjusted, eval=FALSE}
vcov(fit_binomial, type = "leverage", component = "parameters")
vcov(fit_binomial, type = "finite", component = "parameters")
```

# Bootstrap inference

The following code is not evaluated during vignette construction because a
substantive bootstrap can be computationally intensive.

```{r bootstrap-example, eval=FALSE}
boot_binomial <- bootstrap_LuGPLSIM(
  object = fit_binomial,
  B = 500,
  method = "stratified",
  seed = 2026,
  trace = TRUE
)

boot_binomial

V_bootstrap <- vcov(
  fit_binomial,
  type = "bootstrap",
  component = "parameters",
  bootstrap = boot_binomial
)
```

# Poisson example with exposure

For Poisson models, `offset` is a positive exposure vector on its original
scale. The function internally uses its logarithm as the GLM offset.

```{r simulate-poisson}
set.seed(2027)

n_pois <- 250

Z_pois <- cbind(
  z1 = runif(n_pois, 0, 2),
  z2 = runif(n_pois, 0, 2),
  z3 = runif(n_pois, 0, 2)
)

X_pois <- cbind(
  x1 = rnorm(n_pois),
  x2 = rbinom(n_pois, 1, 0.5)
)

exposure <- runif(n_pois, 1, 4)

alpha_pois <- c(1, 1, 1) / sqrt(3)
beta_pois <- c(0.35, -0.25)

u_pois <- as.numeric(Z_pois %*% alpha_pois)
phi_pois <- 0.4 * sin(2 * u_pois)
phi_pois <- phi_pois - mean(phi_pois)

eta_pois <- as.numeric(X_pois %*% beta_pois + phi_pois)
mu_pois <- exposure * exp(eta_pois)
y_pois <- rpois(n_pois, lambda = mu_pois)
```

```{r fit-poisson}
fit_poisson <- LuGPLSIM(
  y = y_pois,
  X = X_pois,
  Z = Z_pois,
  family = poisson(),
  offset = exposure,
  M = 4,
  maxit = 50,
  tol = 1e-6
)
```

```{r predict-poisson}
mu_hat_pois <- predict(
  fit_poisson,
  newX = X_pois,
  newZ = Z_pois,
  type = "response",
  offset = exposure
)

range(mu_hat_pois)
```

# Practical recommendations

1. Standardize continuous columns of \(X\) and \(Z\) when their scales differ substantially.
2. Keep the same coding, column order, and standardization when predicting new observations.
3. Use model-based covariance for routine summaries and bootstrap inference when uncertainty in the index direction is especially important.
4. Avoid extrapolating far beyond the training range of the fitted single index.
5. Run `devtools::test()` and `devtools::check()` before committing package changes.

# Session information

```{r session-info}
sessionInfo()
```
