---
title: "Simulation Study with pgplsim"
author: "Yoon-Bae Jun"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Simulation Study with pgplsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

library(pgplsim)
```

# Purpose

This vignette illustrates a small Monte Carlo study for a generalized
partially linear single-index model. It evaluates bias, empirical standard
deviation, average model-based standard error, root mean squared error,
confidence-interval coverage, convergence, and estimation of the nonlinear
function.

The number of replications is deliberately small so the vignette builds
quickly. Increase it substantially for a publication-quality study.

# Data-generating mechanism

For each observation,

\[
\operatorname{logit}\{P(Y_i=1)\}
=
X_i^\top\beta_0+\phi(Z_i^\top\alpha_0),
\]

where

\[
\alpha_0=\frac{(1,1,1)^\top}{\sqrt{3}},
\qquad
\beta_0=(1,-0.5)^\top,
\qquad
\phi(u)=\sin(2u)-E\{\sin(2u)\}.
\]

The data-generating model contains no separate intercept in the linear
component. Therefore, the formula used below explicitly removes the default
R intercept with `0 +`.

# Simulation functions

The simulation function returns both a data frame for the public formula
interface and the true quantities used to evaluate estimation performance.

```{r simulation-functions}
simulate_binomial_gplsim <- function(n, seed) {
  set.seed(seed)

  dat <- data.frame(
    z1 = runif(n, 0, 2),
    z2 = runif(n, 0, 2),
    z3 = runif(n, 0, 2),
    x1 = rnorm(n),
    x2 = rbinom(n, 1, 0.5)
  )

  alpha_true <- c(1, 1, 1) / sqrt(3)
  beta_true <- c(1, -0.5)

  Z <- as.matrix(
    dat[, c("z1", "z2", "z3")]
  )

  X <- as.matrix(
    dat[, c("x1", "x2")]
  )

  u <- as.numeric(
    Z %*% alpha_true
  )

  phi <- sin(2 * u)
  phi <- phi - mean(phi)

  eta <- as.numeric(
    X %*% beta_true + phi
  )

  probability <- plogis(eta)

  dat$y <- rbinom(
    n,
    size = 1,
    prob = probability
  )

  list(
    data = dat,
    u = u,
    phi = phi,
    alpha_true = alpha_true,
    beta_true = beta_true
  )
}

wald_coverage <- function(
    estimate,
    standard_error,
    truth,
    level = 0.95
) {
  critical_value <- qnorm(
    1 - (1 - level) / 2
  )

  lower <- estimate -
    critical_value * standard_error

  upper <- estimate +
    critical_value * standard_error

  as.numeric(
    lower <= truth &
      truth <= upper
  )
}
```

# One replication

Generate one sample and fit the model using the public formula interface.

```{r one-replication}
dat_one <- simulate_binomial_gplsim(
  n = 300,
  seed = 1001
)

fit_one <- pgplsim(
  y ~ 0 + x1 + x2,
  index = ~ z1 + z2 + z3,
  data = dat_one$data,
  family = binomial(),
  M = 4,
  control = pgplsim_control(
    maxit = 50,
    tol = 1e-6
  )
)

summary(fit_one)
```

The formula `y ~ 0 + x1 + x2` matches the data-generating mechanism exactly:
the linear component contains coefficients for `x1` and `x2`, but no
intercept.

```{r one-replication-estimates}
cbind(
  truth = dat_one$alpha_true,
  estimate = fit_one$alpha_hat,
  model_se = fit_one$se_alpha
)

cbind(
  truth = dat_one$beta_true,
  estimate = fit_one$beta_hat,
  model_se = fit_one$se_beta
)
```

# Monte Carlo loop

The following loop repeats the same fitting procedure on independently
simulated samples.

```{r monte-carlo}
n_rep <- 20
sample_size <- 300

simulation_results <- vector(
  "list",
  n_rep
)

for (r in seq_len(n_rep)) {

  dat_r <- simulate_binomial_gplsim(
    n = sample_size,
    seed = 10000 + r
  )

  fit_r <- tryCatch(
    pgplsim(
      y ~ 0 + x1 + x2,
      index = ~ z1 + z2 + z3,
      data = dat_r$data,
      family = binomial(),
      M = 4,
      control = pgplsim_control(
        maxit = 50,
        tol = 1e-6
      )
    ),
    error = function(e) e
  )

  if (inherits(fit_r, "error")) {
    simulation_results[[r]] <- list(
      success = FALSE,
      message = conditionMessage(fit_r)
    )

    next
  }

  simulation_results[[r]] <- list(
    success = TRUE,
    converged = isTRUE(fit_r$converged),
    alpha_hat = fit_r$alpha_hat,
    beta_hat = fit_r$beta_hat,
    se_alpha = fit_r$se_alpha,
    se_beta = fit_r$se_beta,
    alpha_coverage = wald_coverage(
      fit_r$alpha_hat,
      fit_r$se_alpha,
      dat_r$alpha_true
    ),
    beta_coverage = wald_coverage(
      fit_r$beta_hat,
      fit_r$se_beta,
      dat_r$beta_true
    ),
    lambda = fit_r$lambda,
    iterations = fit_r$iterations
  )
}
```

# Combine successful replications

A replication is retained for the performance summary only if the model fit
completed successfully and reported convergence.

```{r combine-results}
successful <- vapply(
  simulation_results,
  function(x) {
    isTRUE(x$success) &&
      isTRUE(x$converged)
  },
  logical(1)
)

sum(successful)
mean(successful)

successful_results <-
  simulation_results[successful]
```

For this small illustrative simulation, at least one successful replication is
required before the summaries below can be computed.

```{r extract-results}
if (length(successful_results) == 0L) {
  stop(
    "No simulation replications converged successfully."
  )
}

alpha_estimates <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "alpha_hat"
  )
)

beta_estimates <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "beta_hat"
  )
)

alpha_ses <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "se_alpha"
  )
)

beta_ses <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "se_beta"
  )
)

alpha_coverage <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "alpha_coverage"
  )
)

beta_coverage <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "beta_coverage"
  )
)

lambda_values <- vapply(
  successful_results,
  `[[`,
  numeric(1),
  "lambda"
)

iteration_values <- vapply(
  successful_results,
  `[[`,
  integer(1),
  "iterations"
)
```

# Performance metrics

The helper below summarizes Monte Carlo performance for a block of model
parameters.

```{r performance-function}
summarize_parameter_block <- function(
    estimates,
    standard_errors,
    coverage,
    truth,
    parameter_names
) {
  data.frame(
    parameter = parameter_names,
    truth = truth,
    mean_estimate = colMeans(estimates),
    bias = colMeans(estimates) - truth,
    empirical_sd = apply(
      estimates,
      2,
      sd
    ),
    mean_model_se = colMeans(
      standard_errors
    ),
    rmse = sqrt(
      colMeans(
        sweep(
          estimates,
          2,
          truth,
          FUN = "-"
        )^2
      )
    ),
    coverage = colMeans(coverage),
    row.names = NULL,
    check.names = FALSE
  )
}
```

## Index coefficients

```{r performance-alpha}
alpha_summary <-
  summarize_parameter_block(
    estimates = alpha_estimates,
    standard_errors = alpha_ses,
    coverage = alpha_coverage,
    truth = dat_one$alpha_true,
    parameter_names = paste0(
      "alpha",
      seq_along(dat_one$alpha_true)
    )
  )

alpha_summary
```

## Linear coefficients

```{r performance-beta}
beta_summary <-
  summarize_parameter_block(
    estimates = beta_estimates,
    standard_errors = beta_ses,
    coverage = beta_coverage,
    truth = dat_one$beta_true,
    parameter_names = c(
      "beta_x1",
      "beta_x2"
    )
  )

beta_summary
```

# Convergence and smoothing summaries

```{r convergence-summary}
data.frame(
  successful_replications = sum(successful),
  requested_replications = n_rep,
  convergence_rate = mean(successful),
  mean_iterations = mean(iteration_values),
  median_iterations = median(iteration_values),
  mean_lambda = mean(lambda_values),
  median_lambda = median(lambda_values)
)
```

Because this vignette uses only a small number of replications, these summaries
are intended to illustrate the workflow rather than provide definitive
Monte Carlo evidence.

# Estimated nonlinear function

The public prediction interface uses `newdata`. The fitted index and nonlinear
smooth can therefore be obtained directly from the simulated data frame.

```{r nonlinear-function}
u_hat <- predict(
  fit_one,
  newdata = dat_one$data,
  type = "index"
)

smooth_hat <- predict(
  fit_one,
  newdata = dat_one$data,
  type = "smooth"
)

ordering_true <- order(
  dat_one$u
)

ordering_hat <- order(
  u_hat
)

plot(
  dat_one$u[ordering_true],
  dat_one$phi[ordering_true],
  type = "l",
  lwd = 2,
  xlab = "Single index",
  ylab = "Smooth function",
  main = "True and estimated nonlinear functions"
)

lines(
  u_hat[ordering_hat],
  smooth_hat[ordering_hat],
  lty = 2,
  lwd = 2
)

legend(
  "topright",
  legend = c(
    "True",
    "Estimated"
  ),
  lty = c(1, 2),
  lwd = 2,
  bty = "n"
)
```

The fitted index is identifiable only under the package normalization
constraint on \(\alpha\), while the smooth is estimated on the fitted
single-index scale.

# Optional bootstrap comparison

Bootstrap inference is available through `bootstrap_pgplsim()`. The example
below is not evaluated during vignette construction because a substantive
bootstrap may be computationally intensive.

```{r bootstrap-comparison, eval=FALSE}
boot_one <- bootstrap_pgplsim(
  object = fit_one,
  B = 500,
  method = "stratified",
  seed = 2026,
  max_attempts = 800,
  trace = TRUE
)

data.frame(
  parameter = names(fit_one$alpha_hat),
  model_se = fit_one$se_alpha,
  bootstrap_se = boot_one$se_alpha
)
```

# Extending the study

A full simulation study can vary sample size, outcome family,
nonlinear-function shape, index-effect strength, correlation among index
covariates, covariate dimension, spline-basis dimension, and variance
estimator.

For a publication-quality study, increase the number of Monte Carlo
replications substantially and save replication-level output rather than only
the final averages. Replication-level results make it possible to examine
convergence failures, smoothing-parameter behavior, outlying estimates, and
coverage in greater detail.

The same framework can also be extended to Poisson outcomes. Under the current
public API, a Poisson exposure should be supplied as an additive link-scale
offset, for example `offset = log(exposure)`.

# Session information

```{r session-info}
sessionInfo()
```
