---
title: "Simulation Study with pgplsim"
author: "Yoon-Bae Jun"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Simulation Study with pgplsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
library(pgplsim)
```

# Purpose

This vignette illustrates a small Monte Carlo study for a generalized
partially linear single-index model. It evaluates bias, empirical standard
deviation, average model-based standard error, root mean squared error,
confidence-interval coverage, convergence, and estimation of the nonlinear
function.

The number of replications is deliberately small so the vignette builds
quickly. Increase it substantially for a publication-quality study.

# Data-generating mechanism

For each observation,

\[
\operatorname{logit}\{P(Y_i=1)\}
=
X_i^\top\beta_0+\phi(Z_i^\top\alpha_0),
\]

where

\[
\alpha_0=\frac{(1,1,1)^\top}{\sqrt{3}},
\qquad
\beta_0=(1,-0.5)^\top,
\qquad
\phi(u)=\sin(2u)-E\{\sin(2u)\}.
\]

# Simulation functions

```{r simulation-functions}
simulate_binomial_gplsim <- function(n, seed) {
  set.seed(seed)

  Z <- cbind(
    z1 = runif(n, 0, 2),
    z2 = runif(n, 0, 2),
    z3 = runif(n, 0, 2)
  )

  X <- cbind(
    x1 = rnorm(n),
    x2 = rbinom(n, 1, 0.5)
  )

  alpha_true <- c(1, 1, 1) / sqrt(3)
  beta_true <- c(1, -0.5)

  u <- as.numeric(Z %*% alpha_true)
  phi <- sin(2 * u)
  phi <- phi - mean(phi)

  eta <- as.numeric(X %*% beta_true + phi)
  probability <- plogis(eta)
  y <- rbinom(n, size = 1, prob = probability)

  list(
    y = y,
    X = X,
    Z = Z,
    u = u,
    phi = phi,
    alpha_true = alpha_true,
    beta_true = beta_true
  )
}

wald_coverage <- function(estimate, standard_error, truth, level = 0.95) {
  critical_value <- qnorm(1 - (1 - level) / 2)
  lower <- estimate - critical_value * standard_error
  upper <- estimate + critical_value * standard_error
  as.numeric(lower <= truth & truth <= upper)
}
```

# One replication

```{r one-replication}
dat_one <- simulate_binomial_gplsim(
  n = 300,
  seed = 1001
)

fit_one <- LuGPLSIM(
  y = dat_one$y,
  X = dat_one$X,
  Z = dat_one$Z,
  family = binomial(),
  M = 4,
  maxit = 50,
  tol = 1e-6
)

summary(fit_one)
```

```{r one-replication-estimates}
cbind(
  truth = dat_one$alpha_true,
  estimate = fit_one$alpha_hat,
  model_se = fit_one$se_alpha
)

cbind(
  truth = dat_one$beta_true,
  estimate = fit_one$beta_hat,
  model_se = fit_one$se_beta
)
```

# Monte Carlo loop

```{r monte-carlo}
n_rep <- 20
sample_size <- 300
simulation_results <- vector("list", n_rep)

for (r in seq_len(n_rep)) {
  dat_r <- simulate_binomial_gplsim(
    n = sample_size,
    seed = 10000 + r
  )

  fit_r <- tryCatch(
    LuGPLSIM(
      y = dat_r$y,
      X = dat_r$X,
      Z = dat_r$Z,
      family = binomial(),
      M = 4,
      maxit = 50,
      tol = 1e-6
    ),
    error = function(e) e
  )

  if (inherits(fit_r, "error")) {
    simulation_results[[r]] <- list(
      success = FALSE,
      message = conditionMessage(fit_r)
    )
    next
  }

  simulation_results[[r]] <- list(
    success = TRUE,
    converged = isTRUE(fit_r$converged),
    alpha_hat = fit_r$alpha_hat,
    beta_hat = fit_r$beta_hat,
    se_alpha = fit_r$se_alpha,
    se_beta = fit_r$se_beta,
    alpha_coverage = wald_coverage(
      fit_r$alpha_hat,
      fit_r$se_alpha,
      dat_r$alpha_true
    ),
    beta_coverage = wald_coverage(
      fit_r$beta_hat,
      fit_r$se_beta,
      dat_r$beta_true
    ),
    lambda = fit_r$lambda,
    iterations = fit_r$iterations
  )
}
```

# Combine successful replications

```{r combine-results}
successful <- vapply(
  simulation_results,
  function(x) isTRUE(x$success) && isTRUE(x$converged),
  logical(1)
)

sum(successful)
mean(successful)

successful_results <- simulation_results[successful]
```

```{r extract-results}
alpha_estimates <- do.call(
  rbind,
  lapply(successful_results, `[[`, "alpha_hat")
)

beta_estimates <- do.call(
  rbind,
  lapply(successful_results, `[[`, "beta_hat")
)

alpha_ses <- do.call(
  rbind,
  lapply(successful_results, `[[`, "se_alpha")
)

beta_ses <- do.call(
  rbind,
  lapply(successful_results, `[[`, "se_beta")
)

alpha_coverage <- do.call(
  rbind,
  lapply(successful_results, `[[`, "alpha_coverage")
)

beta_coverage <- do.call(
  rbind,
  lapply(successful_results, `[[`, "beta_coverage")
)

lambda_values <- vapply(
  successful_results,
  `[[`,
  numeric(1),
  "lambda"
)

iteration_values <- vapply(
  successful_results,
  `[[`,
  integer(1),
  "iterations"
)
```

# Performance metrics

```{r performance-function}
summarize_parameter_block <- function(
    estimates,
    standard_errors,
    coverage,
    truth,
    parameter_names
) {
  data.frame(
    parameter = parameter_names,
    truth = truth,
    mean_estimate = colMeans(estimates),
    bias = colMeans(estimates) - truth,
    empirical_sd = apply(estimates, 2, sd),
    mean_model_se = colMeans(standard_errors),
    rmse = sqrt(colMeans(
      sweep(estimates, 2, truth, FUN = "-")^2
    )),
    coverage = colMeans(coverage),
    row.names = NULL,
    check.names = FALSE
  )
}
```

```{r performance-alpha}
alpha_summary <- summarize_parameter_block(
  estimates = alpha_estimates,
  standard_errors = alpha_ses,
  coverage = alpha_coverage,
  truth = dat_one$alpha_true,
  parameter_names = paste0("alpha", seq_along(dat_one$alpha_true))
)

alpha_summary
```

```{r performance-beta}
beta_summary <- summarize_parameter_block(
  estimates = beta_estimates,
  standard_errors = beta_ses,
  coverage = beta_coverage,
  truth = dat_one$beta_true,
  parameter_names = paste0("beta", seq_along(dat_one$beta_true))
)

beta_summary
```

# Convergence and smoothing summaries

```{r convergence-summary}
data.frame(
  successful_replications = sum(successful),
  requested_replications = n_rep,
  mean_iterations = mean(iteration_values),
  median_iterations = median(iteration_values),
  mean_lambda = mean(lambda_values),
  median_lambda = median(lambda_values)
)
```

# Estimated nonlinear function

```{r nonlinear-function}
u_hat <- predict(
  fit_one,
  newX = dat_one$X,
  newZ = dat_one$Z,
  type = "index"
)

smooth_hat <- predict(
  fit_one,
  newX = dat_one$X,
  newZ = dat_one$Z,
  type = "smooth"
)

ordering_true <- order(dat_one$u)
ordering_hat <- order(u_hat)

plot(
  dat_one$u[ordering_true],
  dat_one$phi[ordering_true],
  type = "l",
  lwd = 2,
  xlab = "Single index",
  ylab = "Smooth function",
  main = "True and estimated nonlinear functions"
)

lines(
  u_hat[ordering_hat],
  smooth_hat[ordering_hat],
  lty = 2,
  lwd = 2
)

legend(
  "topright",
  legend = c("True", "Estimated"),
  lty = c(1, 2),
  lwd = 2,
  bty = "n"
)
```

# Optional bootstrap comparison

```{r bootstrap-comparison, eval=FALSE}
boot_one <- bootstrap_LuGPLSIM(
  object = fit_one,
  B = 500,
  method = "stratified",
  seed = 2026,
  max_attempts = 800,
  trace = TRUE
)

data.frame(
  parameter = names(fit_one$alpha_hat),
  model_se = fit_one$se_alpha,
  bootstrap_se = boot_one$se_alpha
)
```

# Extending the study

A full study can vary sample size, outcome family, nonlinear-function shape,
index-effect strength, correlation among index covariates, covariate dimension,
number of spline knots, and variance estimator. Save replication-level output
rather than only final averages so convergence and smoothing behavior can be
examined later.

# Session information

```{r session-info}
sessionInfo()
```
