---
title: "NHANES Diabetes Analysis with PSPLINE"
author: "Yoon-Bae Jun"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{NHANES Diabetes Analysis with PSPLINE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

has_nhanes <- requireNamespace("NHANES", quietly = TRUE)
has_mgcv <- requireNamespace("mgcv", quietly = TRUE)
```

# Purpose

This vignette demonstrates a binary-outcome analysis using data distributed
with the optional `NHANES` R package. Diabetes status is modeled using:

- a linear demographic component;
- a nonlinear metabolic single-index component;
- model summaries, predicted probabilities, and comparison models.

This is a reproducible package illustration rather than a definitive
epidemiologic analysis. Population-level NHANES inference should incorporate
survey weights, strata, and primary sampling units. The current `PSPLINE`
implementation does not yet fit complex-survey-weighted models.

# Load the package

```{r load-package}
library(PSPLINE)
```

```{r package-availability}
if (!has_nhanes) {
  cat(
    "The optional NHANES package is not installed. ",
    "Install it with install.packages('NHANES') to run this vignette."
  )
}
```

# Load and inspect NHANES

```{r load-nhanes, eval=has_nhanes}
data("NHANES", package = "NHANES")

dim(NHANES)
names(NHANES)[1:20]
```

The vignette checks that the variables used below are available.

```{r required-variables, eval=has_nhanes}
required_variables <- c(
  "Diabetes",
  "Age",
  "Gender",
  "Race1",
  "BMI",
  "BPSysAve",
  "DirectChol"
)

missing_variables <- setdiff(
  required_variables,
  names(NHANES)
)

if (length(missing_variables) > 0L) {
  stop(
    paste0(
      "The installed NHANES data set is missing: ",
      paste(missing_variables, collapse = ", "),
      "."
    )
  )
}
```

# Construct the analysis sample

```{r prepare-data, eval=has_nhanes}
nhanes_analysis <- NHANES[, required_variables]

nhanes_analysis <- nhanes_analysis[
  stats::complete.cases(nhanes_analysis),
]

nhanes_analysis$diabetes_binary <- ifelse(
  as.character(nhanes_analysis$Diabetes) == "Yes",
  1,
  0
)

table(nhanes_analysis$diabetes_binary)
nrow(nhanes_analysis)
```

# Define the linear and index components

The linear component includes age, gender, and race/ethnicity. The metabolic
single index includes body mass index, average systolic blood pressure, and
direct cholesterol.

```{r design-matrices, eval=has_nhanes}
X_raw <- model.matrix(
  ~ Age + Gender + Race1,
  data = nhanes_analysis
)

# Remove the model-matrix intercept because the current PSPLINE
# formulation uses the supplied columns directly.
X_raw <- X_raw[, -1, drop = FALSE]

Z_raw <- cbind(
  BMI = nhanes_analysis$BMI,
  BPSysAve = nhanes_analysis$BPSysAve,
  DirectChol = nhanes_analysis$DirectChol
)

X <- scale(X_raw)
Z <- scale(Z_raw)
y <- nhanes_analysis$diabetes_binary

dim(X)
dim(Z)
```

Save the training centers and scales for future prediction:

```{r scaling-information, eval=has_nhanes}
X_center <- attr(X, "scaled:center")
X_scale <- attr(X, "scaled:scale")

Z_center <- attr(Z, "scaled:center")
Z_scale <- attr(Z, "scaled:scale")
```

# Fit PSPLINE

```{r fit-pspline, eval=has_nhanes}
fit_nhanes <- LuGPLSIM(
  y = y,
  X = X,
  Z = Z,
  family = binomial(),
  M = 4,
  maxit = 100,
  tol = 1e-6
)

fit_nhanes
summary(fit_nhanes)
```

# Interpret the estimated index

```{r index-coefficients, eval=has_nhanes}
data.frame(
  variable = names(fit_nhanes$alpha_hat),
  alpha = as.numeric(fit_nhanes$alpha_hat),
  model_se = as.numeric(fit_nhanes$se_alpha),
  row.names = NULL
)
```

Because the three index variables were standardized, their coefficient
magnitudes are directly comparable. The sign of an index coefficient must be
interpreted together with the estimated smooth function. The package fixes the
first component of \(\alpha\) to be positive for identifiability.

# Plot the estimated nonlinear association

```{r nonlinear-plot, eval=has_nhanes}
u_hat <- predict(
  fit_nhanes,
  newX = X,
  newZ = Z,
  type = "index"
)

smooth_hat <- predict(
  fit_nhanes,
  newX = X,
  newZ = Z,
  type = "smooth"
)

ordering <- order(u_hat)

plot(
  u_hat[ordering],
  smooth_hat[ordering],
  type = "l",
  lwd = 2,
  xlab = "Estimated metabolic index",
  ylab = "Estimated smooth component",
  main = "Estimated nonlinear diabetes-risk function"
)

rug(u_hat)
```

# Predicted probabilities

```{r predicted-probabilities, eval=has_nhanes}
predicted_probability <- predict(
  fit_nhanes,
  newX = X,
  newZ = Z,
  type = "response"
)

summary(predicted_probability)
range(predicted_probability)
```

A simple apparent classification summary at threshold 0.5 is:

```{r classification, eval=has_nhanes}
predicted_class <- as.integer(
  predicted_probability >= 0.5
)

classification_table <- table(
  observed = y,
  predicted = predicted_class
)

classification_table
mean(predicted_class == y)
```

This is an in-sample summary and will generally be optimistic. Use
cross-validation or external validation for predictive assessment.

# Compare with logistic regression

A conventional logistic model treats the metabolic variables as linear terms.

```{r fit-glm, eval=has_nhanes}
glm_data <- data.frame(
  y = y,
  X,
  Z,
  check.names = FALSE
)

fit_glm <- glm(
  y ~ .,
  data = glm_data,
  family = binomial()
)

summary(fit_glm)
```

```{r compare-loglik, eval=has_nhanes}
data.frame(
  model = c(
    "PSPLINE",
    "Logistic regression"
  ),
  log_likelihood = c(
    fit_nhanes$logLik,
    as.numeric(logLik(fit_glm))
  )
)
```

These values should be interpreted cautiously because the penalized smooth has
effective degrees of freedom that are not represented by a simple raw
parameter count.

# Optional generalized additive model

If `mgcv` is installed, a GAM can estimate separate smooth functions of the
three metabolic variables.

```{r fit-gam, eval=has_nhanes && has_mgcv}
gam_data <- data.frame(
  y = y,
  age = as.numeric(scale(nhanes_analysis$Age)),
  gender = nhanes_analysis$Gender,
  race = nhanes_analysis$Race1,
  bmi = as.numeric(scale(nhanes_analysis$BMI)),
  bp = as.numeric(scale(nhanes_analysis$BPSysAve)),
  cholesterol = as.numeric(scale(nhanes_analysis$DirectChol))
)

fit_gam <- mgcv::gam(
  y ~ age + gender + race +
    s(bmi) + s(bp) + s(cholesterol),
  data = gam_data,
  family = binomial(),
  method = "REML"
)

summary(fit_gam)
```

The models have different interpretations: the GAM estimates three separate
smooth effects, whereas PSPLINE estimates one nonlinear function of a weighted
metabolic index.

# Uncertainty estimation

```{r uncertainty, eval=has_nhanes}
V_model <- vcov(
  fit_nhanes,
  type = "model",
  component = "parameters"
)

V_sandwich <- vcov(
  fit_nhanes,
  type = "sandwich",
  component = "parameters"
)

c(
  model_dimension = nrow(V_model),
  sandwich_dimension = nrow(V_sandwich)
)
```

# Optional bootstrap inference

```{r bootstrap-nhanes, eval=FALSE}
boot_nhanes <- bootstrap_LuGPLSIM(
  object = fit_nhanes,
  B = 500,
  method = "stratified",
  seed = 2026,
  max_attempts = 800,
  trace = TRUE
)

data.frame(
  variable = names(fit_nhanes$alpha_hat),
  estimate = fit_nhanes$alpha_hat,
  model_se = fit_nhanes$se_alpha,
  bootstrap_se = boot_nhanes$se_alpha
)
```

# Prediction for new participants

Use the same factor coding, design-matrix columns, centers, and scales that were
used for the training sample.

```{r new-data-prediction, eval=FALSE}
new_X_raw <- model.matrix(
  ~ Age + Gender + Race1,
  data = new_participants
)[, -1, drop = FALSE]

new_Z_raw <- cbind(
  BMI = new_participants$BMI,
  BPSysAve = new_participants$BPSysAve,
  DirectChol = new_participants$DirectChol
)

new_X <- sweep(new_X_raw, 2, X_center, FUN = "-")
new_X <- sweep(new_X, 2, X_scale, FUN = "/")

new_Z <- sweep(new_Z_raw, 2, Z_center, FUN = "-")
new_Z <- sweep(new_Z, 2, Z_scale, FUN = "/")

new_probability <- predict(
  fit_nhanes,
  newX = new_X,
  newZ = new_Z,
  type = "response"
)
```

Factor levels and design-matrix columns must match the fitted model.

# Limitations

1. The example does not incorporate the NHANES survey design.
2. It uses complete-case analysis.
3. It reports apparent rather than externally validated performance.
4. It does not establish causal effects.
5. It assumes that one smooth function of a single metabolic index adequately
   represents the joint nonlinear association.

# Session information

```{r session-info}
sessionInfo()
```
