---
title: "Introduction to pgplsim"
author: "Yoon-Bae Jun"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to pgplsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

`pgplsim` fits generalized partially linear single-index models of the form

\[
g\{E(Y_i\mid X_i,Z_i)\}
=
X_i^\top\beta+\phi(Z_i^\top\alpha),
\]

where \(X_i\) enters linearly, \(Z_i\) enters through a single index, and
\(\phi(\cdot)\) is an unknown smooth function estimated using penalized
splines. The package normalizes \(\alpha\) so that \(\|\alpha\|=1\) and its
first component is positive.

The current implementation supports binomial and Poisson outcomes.

The public fitting interface follows standard R modeling conventions:

```r
pgplsim(
  formula,
  index,
  data,
  family = binomial(),
  offset = NULL,
  M = NULL,
  lambda = 1,
  control = pgplsim_control()
)
```

Here, `formula` specifies the response and linear component, while `index`
specifies the variables entering the single-index component.

# Installation

The development version can be installed from GitHub with:

```{r installation, eval=FALSE}
remotes::install_github("junpeea/pgplsim")
```

Load the package with:

```{r load-package}
library(pgplsim)
```

# A binomial example

## Simulate data

We first simulate a binary outcome from a partially linear single-index model.

```{r simulate-binomial}
set.seed(2026)

n <- 250

dat_binomial <- data.frame(
  z1 = runif(n, 0, 2),
  z2 = runif(n, 0, 2),
  z3 = runif(n, 0, 2),
  x1 = rnorm(n),
  x2 = rbinom(n, 1, 0.5)
)

alpha_true <- c(1, 1, 1) / sqrt(3)
beta_true <- c(0.8, -0.5)

Z <- as.matrix(dat_binomial[, c("z1", "z2", "z3")])
X <- as.matrix(dat_binomial[, c("x1", "x2")])

u_true <- as.numeric(Z %*% alpha_true)

phi_true <- sin(2 * u_true)
phi_true <- phi_true - mean(phi_true)

eta_true <- as.numeric(
  X %*% beta_true + phi_true
)

probability <- plogis(eta_true)

dat_binomial$y <- rbinom(
  n,
  size = 1,
  prob = probability
)

mean(dat_binomial$y)
```

## Fit the model

The linear component is specified through `formula`, and the single-index
component is specified separately through `index`.

```{r fit-binomial}
fit_binomial <- pgplsim(
  y ~ x1 + x2,
  index = ~ z1 + z2 + z3,
  data = dat_binomial,
  family = binomial(),
  M = 4,
  control = pgplsim_control(
    maxit = 50,
    tol = 1e-6
  )
)
```

The usual print and summary methods are available:

```{r print-and-summary}
fit_binomial
summary(fit_binomial)
```

The fitted index direction is normalized for identifiability:

```{r alpha-constraint}
sum(fit_binomial$alpha_hat^2)
fit_binomial$alpha_hat[1]
```

The first quantity should be close to 1, and the first component of
`alpha_hat` should be positive.

# Prediction components

Predictions follow the standard R `newdata` convention. The package can return
the estimated index, nonlinear smooth component, linear component, full linear
predictor, or response-scale fitted value.

```{r prediction-components}
u_hat <- predict(
  fit_binomial,
  newdata = dat_binomial,
  type = "index"
)

smooth_hat <- predict(
  fit_binomial,
  newdata = dat_binomial,
  type = "smooth"
)

linear_hat <- predict(
  fit_binomial,
  newdata = dat_binomial,
  type = "linear"
)

eta_hat <- predict(
  fit_binomial,
  newdata = dat_binomial,
  type = "link"
)

probability_hat <- predict(
  fit_binomial,
  newdata = dat_binomial,
  type = "response"
)

head(
  cbind(
    u_hat,
    smooth_hat,
    linear_hat,
    eta_hat,
    probability_hat
  )
)

range(probability_hat)
```

For predictions on the original fitting sample, `newdata` may be omitted:

```{r prediction-in-sample}
head(
  predict(
    fit_binomial,
    type = "response"
  )
)
```

The package also retains `newX` and `newZ` as a backward-compatible,
lower-level prediction interface. For new code, `newdata` is recommended.

# Plot the nonlinear component

The fitted nonlinear component can be visualized with the package plot method:

```{r plot-method}
plot(fit_binomial)
```

A direct plot can also be constructed from prediction components:

```{r custom-plot}
ordering <- order(u_hat)

plot(
  u_hat[ordering],
  smooth_hat[ordering],
  type = "l",
  xlab = expression(hat(u) == Z^T * hat(alpha)),
  ylab = expression(hat(phi)(hat(u))),
  main = "Estimated single-index smooth"
)
```

# Variance estimation

The model-based covariance matrix for all model parameters is obtained with:

```{r vcov-model}
V_model <- vcov(
  fit_binomial,
  type = "model",
  component = "parameters"
)

dim(V_model)
round(V_model, 4)
```

Separate covariance blocks are also available:

```{r vcov-blocks}
V_alpha <- vcov(
  fit_binomial,
  type = "model",
  component = "alpha"
)

V_beta <- vcov(
  fit_binomial,
  type = "model",
  component = "beta"
)

dim(V_alpha)
dim(V_beta)
```

A sandwich covariance estimator is obtained with:

```{r vcov-sandwich}
V_sandwich <- vcov(
  fit_binomial,
  type = "sandwich",
  component = "parameters"
)

dim(V_sandwich)
```

Leverage-adjusted and finite-sample-corrected covariance estimators are also
available:

```{r vcov-adjusted, eval=FALSE}
vcov(
  fit_binomial,
  type = "leverage",
  component = "parameters"
)

vcov(
  fit_binomial,
  type = "finite",
  component = "parameters"
)
```

# Bootstrap inference

Bootstrap inference is available through `bootstrap_pgplsim()`.

The following example is not evaluated during vignette construction because a
substantive bootstrap can be computationally intensive.

```{r bootstrap-example, eval=FALSE}
boot_binomial <- bootstrap_pgplsim(
  object = fit_binomial,
  B = 500,
  method = "stratified",
  seed = 2026,
  trace = TRUE
)

boot_binomial

V_bootstrap <- vcov(
  fit_binomial,
  type = "bootstrap",
  component = "parameters",
  bootstrap = boot_binomial
)
```

# Poisson example with exposure

For Poisson models, the public `offset` argument is an **additive offset on the
link scale**, following the usual GLM convention.

Therefore, if \(E_i\) is a positive exposure such as population,
person-time, or follow-up time, supply

\[
\log(E_i)
\]

as the offset.

## Simulate Poisson data

```{r simulate-poisson}
set.seed(2027)

n_pois <- 250

dat_poisson <- data.frame(
  z1 = runif(n_pois, 0, 2),
  z2 = runif(n_pois, 0, 2),
  z3 = runif(n_pois, 0, 2),
  x1 = rnorm(n_pois),
  x2 = rbinom(n_pois, 1, 0.5),
  exposure = runif(n_pois, 1, 4)
)

alpha_pois <- c(1, 1, 1) / sqrt(3)
beta_pois <- c(0.35, -0.25)

Z_pois <- as.matrix(
  dat_poisson[, c("z1", "z2", "z3")]
)

X_pois <- as.matrix(
  dat_poisson[, c("x1", "x2")]
)

u_pois <- as.numeric(
  Z_pois %*% alpha_pois
)

phi_pois <- 0.4 * sin(2 * u_pois)
phi_pois <- phi_pois - mean(phi_pois)

eta_pois <- as.numeric(
  X_pois %*% beta_pois + phi_pois
)

mu_pois <- dat_poisson$exposure * exp(eta_pois)

dat_poisson$y <- rpois(
  n_pois,
  lambda = mu_pois
)
```

## Fit the Poisson model

```{r fit-poisson}
fit_poisson <- pgplsim(
  y ~ x1 + x2,
  index = ~ z1 + z2 + z3,
  data = dat_poisson,
  family = poisson(),
  offset = log(dat_poisson$exposure),
  M = 4,
  control = pgplsim_control(
    maxit = 50,
    tol = 1e-6
  )
)
```

## Predict Poisson means

For in-sample prediction, the fitted offset is used automatically when
`offset` is omitted:

```{r predict-poisson-in-sample}
mu_hat_pois <- predict(
  fit_poisson,
  type = "response"
)

range(mu_hat_pois)
```

For prediction using `newdata`, supply the corresponding link-scale offset:

```{r predict-poisson-newdata}
mu_hat_pois_new <- predict(
  fit_poisson,
  newdata = dat_poisson,
  type = "response",
  offset = log(dat_poisson$exposure)
)

range(mu_hat_pois_new)
```

A scalar offset is also allowed and is recycled across observations. For
example, `offset = 0` corresponds to exposure equal to 1.

# Practical recommendations

1. Standardize continuous predictors when their scales differ substantially,
   particularly for variables entering the single-index component.
2. Use ordinary R formula syntax for the linear component and a one-sided
   formula for the single-index component.
3. Keep factor coding and continuous-variable transformations consistent
   between the training data and `newdata`.
4. For Poisson models, remember that the public `offset` is supplied on the
   link scale; use `log(exposure)` for a positive exposure variable.
5. Use model-based covariance for routine summaries and bootstrap inference
   when uncertainty in the estimated index direction is especially important.
6. Avoid extrapolating far beyond the range of the fitted single index.
7. Run `devtools::test()` and `devtools::check()` before committing package
   changes.

# Session information

```{r session-info}
sessionInfo()
```
