---
title: "NHANES Diabetes Analysis with pgplsim"
author: "Yoon-Bae Jun"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{NHANES Diabetes Analysis with pgplsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

has_nhanes <- requireNamespace(
  "NHANES",
  quietly = TRUE
)

has_mgcv <- requireNamespace(
  "mgcv",
  quietly = TRUE
)
```

# Purpose

This vignette demonstrates a binary-outcome analysis using data distributed
with the optional `NHANES` R package. Diabetes status is modeled using:

- a linear demographic component;
- a nonlinear metabolic single-index component;
- model summaries and predicted probabilities;
- comparison with conventional logistic regression and, optionally, a
  generalized additive model.

This is a reproducible package illustration rather than a definitive
epidemiologic analysis. Population-level NHANES inference should incorporate
survey weights, strata, and primary sampling units. The current `pgplsim`
implementation does not fit complex-survey-weighted models.

# Load the package

```{r load-package}
library(pgplsim)
```

```{r package-availability}
if (!has_nhanes) {
  cat(
    "The optional NHANES package is not installed. ",
    "Install it with install.packages('NHANES') ",
    "to run the NHANES-specific code in this vignette."
  )
}
```

# Load and inspect NHANES

```{r load-nhanes, eval=has_nhanes}
data(
  "NHANES",
  package = "NHANES"
)

dim(NHANES)
names(NHANES)[1:20]
```

The vignette checks that the variables used below are available.

```{r required-variables, eval=has_nhanes}
required_variables <- c(
  "Diabetes",
  "Age",
  "Gender",
  "Race1",
  "BMI",
  "BPSysAve",
  "DirectChol"
)

missing_variables <- setdiff(
  required_variables,
  names(NHANES)
)

if (length(missing_variables) > 0L) {
  stop(
    paste0(
      "The installed NHANES data set is missing: ",
      paste(
        missing_variables,
        collapse = ", "
      ),
      "."
    )
  )
}
```

# Construct the analysis sample

We use complete observations for the variables required in this illustration.

```{r prepare-data, eval=has_nhanes}
nhanes_analysis <- NHANES[
  ,
  required_variables
]

nhanes_analysis <- nhanes_analysis[
  stats::complete.cases(nhanes_analysis),
  ,
  drop = FALSE
]

nhanes_analysis <- nhanes_analysis[
  as.character(nhanes_analysis$Diabetes) %in%
    c("Yes", "No"),
  ,
  drop = FALSE
]

nhanes_analysis$diabetes_binary <- as.integer(
  as.character(nhanes_analysis$Diabetes) == "Yes"
)

table(
  nhanes_analysis$Diabetes,
  nhanes_analysis$diabetes_binary
)

nrow(nhanes_analysis)
```

# Define the linear and index components

The linear component contains age, gender, and race/ethnicity:

\[
X_i^\top\beta
=
\beta_0+
\beta_1\operatorname{Age}_i+
\text{gender terms}+
\text{race/ethnicity terms}.
\]

The nonlinear component is based on a single metabolic index constructed from:

- body mass index (`BMI`);
- average systolic blood pressure (`BPSysAve`);
- direct cholesterol (`DirectChol`).

Because the scale of the index variables affects the interpretation of the
estimated index coefficients, we standardize these three continuous variables
before fitting the model.

The formula interface handles the demographic factor coding automatically.

## Standardize the index variables

```{r standardize-index, eval=has_nhanes}
index_variables <- c(
  "BMI",
  "BPSysAve",
  "DirectChol"
)

index_center <- vapply(
  nhanes_analysis[index_variables],
  mean,
  numeric(1)
)

index_scale <- vapply(
  nhanes_analysis[index_variables],
  sd,
  numeric(1)
)

if (any(!is.finite(index_scale) | index_scale <= 0)) {
  stop(
    "All index variables must have positive finite standard deviations."
  )
}

nhanes_analysis$BMI_z <- (
  nhanes_analysis$BMI -
    index_center["BMI"]
) / index_scale["BMI"]

nhanes_analysis$BPSysAve_z <- (
  nhanes_analysis$BPSysAve -
    index_center["BPSysAve"]
) / index_scale["BPSysAve"]

nhanes_analysis$DirectChol_z <- (
  nhanes_analysis$DirectChol -
    index_center["DirectChol"]
) / index_scale["DirectChol"]

summary(
  nhanes_analysis[
    c(
      "BMI_z",
      "BPSysAve_z",
      "DirectChol_z"
    )
  ]
)
```

The saved `index_center` and `index_scale` values are also used when predicting
for new participants.

# Fit pgplsim

The demographic variables enter through the usual two-sided R formula. The
standardized metabolic variables enter through the one-sided `index` formula.

```{r fit-pgplsim, eval=has_nhanes}
fit_nhanes <- pgplsim(
  diabetes_binary ~ Age + Gender + Race1,
  index = ~ BMI_z + BPSysAve_z + DirectChol_z,
  data = nhanes_analysis,
  family = binomial(),
  M = 4,
  control = pgplsim_control(
    maxit = 100,
    tol = 1e-6
  )
)

fit_nhanes
summary(fit_nhanes)
```

The linear component follows ordinary R formula conventions, including an
intercept unless it is explicitly removed. The single-index component does
not include an intercept.

# Interpret the estimated index

```{r index-coefficients, eval=has_nhanes}
data.frame(
  variable = names(
    fit_nhanes$alpha_hat
  ),
  alpha = as.numeric(
    fit_nhanes$alpha_hat
  ),
  model_se = as.numeric(
    fit_nhanes$se_alpha
  ),
  row.names = NULL
)
```

Because the three index variables were standardized, their coefficient
magnitudes are more directly comparable than they would be on their original
measurement scales.

The sign and magnitude of an index coefficient should not be interpreted as a
stand-alone regression effect. Each coefficient determines how its variable
contributes to the estimated single index

\[
u_i = Z_i^\top\alpha,
\]

and the association with diabetes risk is then determined through the
estimated nonlinear function \(\phi(u_i)\).

For identifiability, `pgplsim` normalizes \(\alpha\) to have unit Euclidean
norm and constrains its first component to be positive.

```{r index-constraint, eval=has_nhanes}
sum(
  fit_nhanes$alpha_hat^2
)

fit_nhanes$alpha_hat[1]
```

Because the index vector is constrained for identifiability, its components
are not unconstrained regression coefficients. The reported componentwise
model-based standard errors should therefore be interpreted together with the
unit-norm and sign constraints rather than as inference for unrelated
Euclidean parameters. Bootstrap inference is illustrated later as an
alternative uncertainty assessment.

# Plot the estimated nonlinear association

The public prediction interface accepts the original modeling data through
`newdata`.

```{r nonlinear-plot, eval=has_nhanes}
u_hat <- predict(
  fit_nhanes,
  newdata = nhanes_analysis,
  type = "index"
)

smooth_hat <- predict(
  fit_nhanes,
  newdata = nhanes_analysis,
  type = "smooth"
)

ordering <- order(u_hat)

plot(
  u_hat[ordering],
  smooth_hat[ordering],
  type = "l",
  lwd = 2,
  xlab = "Estimated metabolic index",
  ylab = "Estimated smooth component",
  main = "Estimated nonlinear diabetes-risk function"
)

rug(u_hat)
```

The vertical scale is the estimated contribution of the nonlinear
single-index component to the log-odds of diabetes. Interpretation should
therefore focus on the shape of the fitted smooth rather than on the sign of
an individual index coefficient alone.

# Predicted probabilities

Response-scale predictions from a binomial model are estimated probabilities.

```{r predicted-probabilities, eval=has_nhanes}
predicted_probability <- predict(
  fit_nhanes,
  newdata = nhanes_analysis,
  type = "response"
)

summary(
  predicted_probability
)

range(
  predicted_probability
)
```

For the original fitting sample, `newdata` may also be omitted:

```{r predicted-probabilities-in-sample, eval=has_nhanes}
predicted_probability_in_sample <- predict(
  fit_nhanes,
  type = "response"
)

max(
  abs(
    predicted_probability -
      predicted_probability_in_sample
  )
)
```

The two prediction calls should agree up to numerical precision.

# Apparent classification summary

A simple in-sample classification summary at a probability threshold of 0.5
is:

```{r classification, eval=has_nhanes}
predicted_class <- as.integer(
  predicted_probability >= 0.5
)

classification_table <- table(
  observed = nhanes_analysis$diabetes_binary,
  predicted = predicted_class
)

classification_table

mean(
  predicted_class ==
    nhanes_analysis$diabetes_binary
)
```

This is an apparent, in-sample summary and will generally be optimistic.
Cross-validation or external validation should be used for formal predictive
assessment. The 0.5 threshold is used only for illustration and is not
presented as an optimal clinical or epidemiologic decision threshold.

# Compare with logistic regression

A conventional logistic regression can treat all three metabolic variables as
separate linear predictors.

Using the same standardized metabolic variables allows a more transparent
comparison with the single-index model.

```{r fit-glm, eval=has_nhanes}
fit_glm <- glm(
  diabetes_binary ~
    Age +
    Gender +
    Race1 +
    BMI_z +
    BPSysAve_z +
    DirectChol_z,
  data = nhanes_analysis,
  family = binomial()
)

summary(fit_glm)
```

## Descriptive log-likelihood values

For reference, the fitted log-likelihood values are:

```{r compare-loglik, eval=has_nhanes}
data.frame(
  model = c(
    "pgplsim",
    "Logistic regression"
  ),
  log_likelihood = c(
    fit_nhanes$logLik,
    as.numeric(
      logLik(fit_glm)
    )
  )
)
```

These values are reported descriptively and should not be treated as a formal
model-selection comparison. The `pgplsim` fit contains a penalized nonlinear
component whose effective complexity is not represented by an ordinary raw
parameter count. Formal predictive comparison would require an appropriate
out-of-sample or cross-validation procedure.

# Optional generalized additive model

If `mgcv` is installed, a generalized additive model can estimate separate
smooth functions for the three metabolic variables.

```{r fit-gam, eval=has_nhanes && has_mgcv}
fit_gam <- mgcv::gam(
  diabetes_binary ~
    Age +
    Gender +
    Race1 +
    s(BMI_z) +
    s(BPSysAve_z) +
    s(DirectChol_z),
  data = nhanes_analysis,
  family = binomial(),
  method = "REML"
)

summary(fit_gam)
```

The models have different interpretations. The GAM estimates three separate
smooth effects, whereas `pgplsim` estimates one nonlinear function of a
weighted metabolic index.

# Uncertainty estimation

Model-based and sandwich covariance estimators are available through `vcov()`.

```{r uncertainty, eval=has_nhanes}
V_model <- vcov(
  fit_nhanes,
  type = "model",
  component = "parameters"
)

V_sandwich <- vcov(
  fit_nhanes,
  type = "sandwich",
  component = "parameters"
)

c(
  model_dimension = nrow(V_model),
  sandwich_dimension = nrow(V_sandwich)
)
```

Separate covariance blocks can also be requested:

```{r uncertainty-components, eval=has_nhanes}
V_alpha <- vcov(
  fit_nhanes,
  type = "model",
  component = "alpha"
)

V_beta <- vcov(
  fit_nhanes,
  type = "model",
  component = "beta"
)

dim(V_alpha)
dim(V_beta)
```

# Optional bootstrap inference

Bootstrap inference is available through `bootstrap_pgplsim()`. The example
below is not evaluated during vignette construction because bootstrap
inference can be computationally intensive. The choice `B = 500` illustrates
a substantive analysis; in applications, the number of bootstrap replications
should be chosen to provide adequate Monte Carlo precision for the inferential
quantities of interest.

```{r bootstrap-nhanes, eval=FALSE}
boot_nhanes <- bootstrap_pgplsim(
  object = fit_nhanes,
  B = 500,
  method = "stratified",
  seed = 2026,
  max_attempts = 800,
  trace = TRUE
)

data.frame(
  variable = names(
    fit_nhanes$alpha_hat
  ),
  estimate = fit_nhanes$alpha_hat,
  model_se = fit_nhanes$se_alpha,
  bootstrap_se = boot_nhanes$se_alpha
)
```

# Prediction for new participants

The formula-based prediction interface removes the need to construct `X` and
`Z` matrices manually. The new data should contain the variables referenced by
the fitted formulas and use compatible factor levels.

The standardized index variables must be constructed using the **training**
centers and scales, not values recomputed from the new participants.

```{r new-data-prediction, eval=FALSE}
new_participants$BMI_z <- (
  new_participants$BMI -
    index_center["BMI"]
) / index_scale["BMI"]

new_participants$BPSysAve_z <- (
  new_participants$BPSysAve -
    index_center["BPSysAve"]
) / index_scale["BPSysAve"]

new_participants$DirectChol_z <- (
  new_participants$DirectChol -
    index_center["DirectChol"]
) / index_scale["DirectChol"]

new_probability <- predict(
  fit_nhanes,
  newdata = new_participants,
  type = "response"
)

new_probability
```

The factor levels for `Gender` and `Race1` must be compatible with those used
when fitting the model. Prediction at previously unseen factor levels is not
defined by the fitted design matrix.

# Practical interpretation

This example separates two types of covariate structure.

The demographic variables enter linearly through

\[
X_i^\top\beta,
\]

so their coefficients have the usual conditional log-odds interpretation
within the fitted model.

The three metabolic variables are reduced to

\[
u_i =
\alpha_1\operatorname{BMI}_{z,i}
+
\alpha_2\operatorname{BPSysAve}_{z,i}
+
\alpha_3\operatorname{DirectChol}_{z,i},
\]

and diabetes risk is associated with this index through the estimated smooth

\[
\phi(u_i).
\]

Consequently, the metabolic index weights and the nonlinear smooth should be
interpreted jointly.

# Limitations

1. The example does not incorporate NHANES survey weights, strata, or primary
   sampling units and therefore should not be interpreted as
   population-representative NHANES inference.
2. The example uses complete-case analysis.
3. The classification summary is apparent rather than externally validated.
4. The analysis is associational and does not establish causal effects.
5. The model assumes that one smooth function of a single metabolic index
   adequately represents the joint nonlinear association of BMI, systolic
   blood pressure, and cholesterol with diabetes status.
6. The metabolic index variables are standardized using this analysis sample,
   so the resulting index is sample-dependent.

# Session information

```{r session-info}
sessionInfo()
```
