---
title: "Simulation Study with pgplsim"
author: "Yoon-Bae Jun"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Simulation Study with pgplsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

library(pgplsim)
```

# Purpose

This vignette illustrates how `pgplsim` can be evaluated in a small Monte
Carlo experiment. The workflow computes bias, empirical standard deviation,
average model-based standard error, root mean squared error, Wald
confidence-interval coverage, convergence, and recovery of the nonlinear
component.

The number of replications is deliberately small so that the vignette builds
quickly. The resulting Monte Carlo summaries are illustrative only and should
not be interpreted as a formal evaluation of the statistical properties of
the estimator. A publication-quality simulation study should use substantially
more replications and examine multiple data-generating scenarios.

# Data-generating mechanism

For each observation,

\[
\operatorname{logit}\{P(Y_i=1)\}
=
X_i^\top\beta_0+\phi(Z_i^\top\alpha_0),
\]

where

\[
\alpha_0=\frac{(1,1,1)^\top}{\sqrt{3}},
\qquad
\beta_0=(1,-0.5)^\top,
\]

with the nonlinear component generated as

\[
\phi_i
=
\sin(2u_i)
-
\frac{1}{n}\sum_{j=1}^{n}\sin(2u_j),
\qquad
u_i=Z_i^\top\alpha_0.
\]

The data-generating model contains no separate intercept in the linear
component. Therefore, the formula used below explicitly removes the default
R intercept with `0 +`. The sample centering of the nonlinear component
matches the no-intercept parameterization used in this illustrative
simulation.

# Simulation functions

The simulation function returns both a data frame for the public formula
interface and the true quantities used to evaluate estimation performance.

```{r simulation-functions}
simulate_binomial_gplsim <- function(n, seed) {
  set.seed(seed)

  dat <- data.frame(
    z1 = runif(n, 0, 2),
    z2 = runif(n, 0, 2),
    z3 = runif(n, 0, 2),
    x1 = rnorm(n),
    x2 = rbinom(n, 1, 0.5)
  )

  alpha_true <- c(1, 1, 1) / sqrt(3)
  beta_true <- c(1, -0.5)

  Z <- as.matrix(
    dat[, c("z1", "z2", "z3")]
  )

  X <- as.matrix(
    dat[, c("x1", "x2")]
  )

  u <- as.numeric(
    Z %*% alpha_true
  )

  phi <- sin(2 * u)
  phi <- phi - mean(phi)

  eta <- as.numeric(
    X %*% beta_true + phi
  )

  probability <- plogis(eta)

  dat$y <- rbinom(
    n,
    size = 1,
    prob = probability
  )

  list(
    data = dat,
    u = u,
    phi = phi,
    alpha_true = alpha_true,
    beta_true = beta_true
  )
}

wald_coverage <- function(
    estimate,
    standard_error,
    truth,
    level = 0.95
) {
  critical_value <- qnorm(
    1 - (1 - level) / 2
  )

  lower <- estimate -
    critical_value * standard_error

  upper <- estimate +
    critical_value * standard_error

  as.numeric(
    lower <= truth &
      truth <= upper
  )
}
```

# One replication

Generate one sample and fit the model using the public formula interface.

```{r one-replication}
dat_one <- simulate_binomial_gplsim(
  n = 300,
  seed = 1001
)

fit_one <- pgplsim(
  y ~ 0 + x1 + x2,
  index = ~ z1 + z2 + z3,
  data = dat_one$data,
  family = binomial(),
  M = 4,
  control = pgplsim_control(
    maxit = 50,
    tol = 1e-6
  )
)

summary(fit_one)
```

The formula `y ~ 0 + x1 + x2` matches the data-generating mechanism exactly:
the linear component contains coefficients for `x1` and `x2`, but no
intercept.

```{r one-replication-estimates}
cbind(
  truth = dat_one$alpha_true,
  estimate = fit_one$alpha_hat,
  model_se = fit_one$se_alpha
)

cbind(
  truth = dat_one$beta_true,
  estimate = fit_one$beta_hat,
  model_se = fit_one$se_beta
)
```

# Monte Carlo loop

The following loop repeats the same fitting procedure on independently
simulated samples. Fitting errors and completed fits that fail to report
convergence are retained separately so that the two types of failure can be
distinguished.

```{r monte-carlo}
n_rep <- 20
sample_size <- 300

simulation_results <- vector(
  "list",
  n_rep
)

for (r in seq_len(n_rep)) {

  dat_r <- simulate_binomial_gplsim(
    n = sample_size,
    seed = 10000 + r
  )

  fit_r <- tryCatch(
    pgplsim(
      y ~ 0 + x1 + x2,
      index = ~ z1 + z2 + z3,
      data = dat_r$data,
      family = binomial(),
      M = 4,
      control = pgplsim_control(
        maxit = 50,
        tol = 1e-6
      )
    ),
    error = function(e) e
  )

  if (inherits(fit_r, "error")) {
    simulation_results[[r]] <- list(
      success = FALSE,
      message = conditionMessage(fit_r)
    )

    next
  }

  simulation_results[[r]] <- list(
    success = TRUE,
    converged = isTRUE(fit_r$converged),
    alpha_hat = fit_r$alpha_hat,
    beta_hat = fit_r$beta_hat,
    se_alpha = fit_r$se_alpha,
    se_beta = fit_r$se_beta,
    alpha_coverage = wald_coverage(
      fit_r$alpha_hat,
      fit_r$se_alpha,
      dat_r$alpha_true
    ),
    beta_coverage = wald_coverage(
      fit_r$beta_hat,
      fit_r$se_beta,
      dat_r$beta_true
    ),
    lambda = fit_r$lambda,
    iterations = fit_r$iterations
  )
}
```

# Combine successful replications

A replication is retained for the parameter-performance summaries only if the
model fit completed successfully and reported convergence. Fitting errors and
nonconverged completed fits are nevertheless counted separately.

```{r combine-results}
fit_completed <- vapply(
  simulation_results,
  function(x) isTRUE(x$success),
  logical(1)
)

converged <- vapply(
  simulation_results,
  function(x) {
    isTRUE(x$success) &&
      isTRUE(x$converged)
  },
  logical(1)
)

successful <- converged

successful_results <-
  simulation_results[successful]

data.frame(
  requested_replications = n_rep,
  completed_fits = sum(fit_completed),
  converged_fits = sum(converged),
  fitting_errors = sum(!fit_completed),
  nonconverged_fits = sum(
    fit_completed & !converged
  ),
  convergence_rate = mean(converged)
)
```

For this small illustrative simulation, at least one successful replication is
required before the summaries below can be computed.

```{r extract-results}
if (length(successful_results) == 0L) {
  stop(
    "No simulation replications converged successfully."
  )
}

alpha_estimates <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "alpha_hat"
  )
)

beta_estimates <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "beta_hat"
  )
)

alpha_ses <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "se_alpha"
  )
)

beta_ses <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "se_beta"
  )
)

alpha_coverage <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "alpha_coverage"
  )
)

beta_coverage <- do.call(
  rbind,
  lapply(
    successful_results,
    `[[`,
    "beta_coverage"
  )
)

lambda_values <- vapply(
  successful_results,
  `[[`,
  numeric(1),
  "lambda"
)

iteration_values <- vapply(
  successful_results,
  `[[`,
  integer(1),
  "iterations"
)
```

# Performance metrics

The helper below summarizes Monte Carlo performance for a block of model
parameters.

```{r performance-function}
summarize_parameter_block <- function(
    estimates,
    standard_errors,
    coverage,
    truth,
    parameter_names
) {
  data.frame(
    parameter = parameter_names,
    truth = truth,
    mean_estimate = colMeans(estimates),
    bias = colMeans(estimates) - truth,
    empirical_sd = apply(
      estimates,
      2,
      sd
    ),
    mean_model_se = colMeans(
      standard_errors
    ),
    rmse = sqrt(
      colMeans(
        sweep(
          estimates,
          2,
          truth,
          FUN = "-"
        )^2
      )
    ),
    coverage = colMeans(coverage),
    row.names = NULL,
    check.names = FALSE
  )
}
```

## A note on inference for the index direction

The index vector \(\alpha\) is subject to the identifiability constraints
\(\|\alpha\|=1\) and \(\alpha_1>0\). Consequently, its components are not
unconstrained Euclidean parameters. The componentwise model-based standard
errors and Wald coverage summaries below describe the inferential quantities
returned by the current implementation, but they should be interpreted with
the normalization constraint in mind.

In particular, the small simulation below is intended to illustrate how these
quantities can be summarized rather than to establish the finite-sample
coverage properties of componentwise Wald intervals for \(\alpha\). Bootstrap
inference provides an alternative approach and is illustrated later without
being executed during vignette construction.

## Index coefficients

```{r performance-alpha}
alpha_summary <-
  summarize_parameter_block(
    estimates = alpha_estimates,
    standard_errors = alpha_ses,
    coverage = alpha_coverage,
    truth = dat_one$alpha_true,
    parameter_names = paste0(
      "alpha",
      seq_along(dat_one$alpha_true)
    )
  )

alpha_summary
```

Because \(\alpha\) represents an index direction, it is also useful to assess
recovery of the vector as a whole rather than only its individual components.
Under the package sign constraint, angular error provides a direct measure of
the discrepancy between the true and estimated index directions.

```{r performance-alpha-direction}
alpha_true <- dat_one$alpha_true

cosine_similarity <- as.numeric(
  alpha_estimates %*% alpha_true
)

cosine_similarity <- pmin(
  1,
  pmax(-1, cosine_similarity)
)

angular_error_deg <- acos(
  cosine_similarity
) * 180 / pi

data.frame(
  mean_angular_error_deg =
    mean(angular_error_deg),
  median_angular_error_deg =
    median(angular_error_deg),
  rmse_alpha_vector = sqrt(
    mean(
      rowSums(
        sweep(
          alpha_estimates,
          2,
          alpha_true,
          FUN = "-"
        )^2
      )
    )
  )
)
```

An angular error of zero degrees corresponds to perfect recovery of the true
index direction. These summaries remain illustrative because only a small
number of Monte Carlo replications is used.

## Linear coefficients

```{r performance-beta}
beta_summary <-
  summarize_parameter_block(
    estimates = beta_estimates,
    standard_errors = beta_ses,
    coverage = beta_coverage,
    truth = dat_one$beta_true,
    parameter_names = c(
      "beta_x1",
      "beta_x2"
    )
  )

beta_summary
```

# Convergence and smoothing summaries

The convergence summary distinguishes fitting errors from fits that completed
but did not satisfy the package convergence criterion.

```{r convergence-summary}
data.frame(
  requested_replications = n_rep,
  completed_fits = sum(fit_completed),
  converged_fits = sum(converged),
  fitting_errors = sum(!fit_completed),
  nonconverged_fits = sum(
    fit_completed & !converged
  ),
  convergence_rate = mean(converged),
  mean_iterations = mean(iteration_values),
  median_iterations = median(iteration_values),
  mean_lambda = mean(lambda_values),
  median_lambda = median(lambda_values)
)
```

Because this vignette uses only a small number of replications, these summaries
are intended to illustrate the workflow rather than provide definitive Monte
Carlo evidence.

# Estimated nonlinear function

The public prediction interface uses `newdata`. The fitted index and nonlinear
smooth can therefore be obtained directly from the simulated data frame.

```{r nonlinear-function}
u_hat <- predict(
  fit_one,
  newdata = dat_one$data,
  type = "index"
)

smooth_hat <- predict(
  fit_one,
  newdata = dat_one$data,
  type = "smooth"
)

ordering_true <- order(
  dat_one$u
)

ordering_hat <- order(
  u_hat
)

plot(
  dat_one$u[ordering_true],
  dat_one$phi[ordering_true],
  type = "l",
  lwd = 2,
  xlab = "Single index",
  ylab = "Smooth function",
  main = "True and estimated nonlinear functions"
)

lines(
  u_hat[ordering_hat],
  smooth_hat[ordering_hat],
  lty = 2,
  lwd = 2
)

legend(
  "topright",
  legend = c(
    "True",
    "Estimated"
  ),
  lty = c(1, 2),
  lwd = 2,
  bty = "n"
)
```

The two curves are shown on their respective true and estimated single-index
scales. Therefore, discrepancies between them can reflect both estimation of
the index direction \(\alpha\) and estimation of the smooth function
\(\phi(\cdot)\). This figure should be interpreted as recovery of the combined
single-index structure rather than as an isolated measure of smooth-function
estimation error.

# Optional bootstrap comparison

Bootstrap inference is available through `bootstrap_pgplsim()`. The example
below is not evaluated during vignette construction because bootstrap
inference can be computationally intensive. The choice `B = 500` illustrates
a substantive analysis; in applications, the number of bootstrap replications
should be chosen to provide adequate Monte Carlo precision for the inferential
quantities of interest.

```{r bootstrap-comparison, eval=FALSE}
boot_one <- bootstrap_pgplsim(
  object = fit_one,
  B = 500,
  method = "stratified",
  seed = 2026,
  max_attempts = 800,
  trace = TRUE
)

data.frame(
  parameter = names(fit_one$alpha_hat),
  model_se = fit_one$se_alpha,
  bootstrap_se = boot_one$se_alpha
)
```

# Extending the study

A full simulation study can vary sample size, outcome family,
nonlinear-function shape, index-effect strength, correlation among index
covariates, covariate dimension, spline-basis dimension, and variance
estimator.

For a publication-quality study, increase the number of Monte Carlo
replications substantially and save replication-level output rather than only
the final averages. Replication-level results make it possible to examine
convergence failures, smoothing-parameter behavior, outlying estimates, and
coverage in greater detail.

The same framework can also be extended to Poisson outcomes. Under the current
public API, a Poisson exposure should be supplied as an additive link-scale
offset, for example `offset = log(exposure)`.

# Session information

```{r session-info}
sessionInfo()
```
